---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# AlzhMarkers

## Overview

The `AlzheiMarkers` package provides tools to analyze metabolomic biomarkers 
for distinguishing Alzheimer's Disease (AD) patients from Healthy Controls (HC).
It offers functionalities for data preparation, machine learning model 
training, model evaluation, and visualization of results, making it a toolkit 
for biomarker analysis in Alzheimer's research.

## Key features

-   **Data preparation**: Functions to load, preprocess, and split data
into training and test sets.
-   **Model training**: Supports multiple machine learning models including
Random Forest, Lasso, XGBoost, Naive Bayes, and Partial Least Squares (PLS).
-   **Evaluation**: Model performance evaluation using ROC curves 
and cross-validation.
-   **Visualization**: Tools for plotting ROC curves and variable importance
across different models.

## Basic usage

```{r eval=FALSE}
library(AlzheiMarkers)
data <- prepare_data("path/to/data.csv") # Prepare data
split_data <- split_data(data)

models <- train_models(split_data$train) # Train models

eval_results <- evaluate_models(models) # Evaluate models
plot_roc_curves(eval_results$eval_results)
```
