---
title: "AlzheiMarkers: An R Package for Metabolomic Biomarker Discovery and
Classification of Alzheimer's Disease Using Machine Learning"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{AlzheiMarkers: An R Package for Metabolomic Biomarker
  Discovery and Classification of Alzheimer's Disease Using Machine Learning}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Abstract

Alzheimer’s disease (AD) is a progressive neurodegenerative disorder
that affects millions of people worldwide. Early diagnosis and
differentiation from healthy aging remain critical challenges in the
field. Metabolomic biomarkers present a promising avenue for addressing
these challenges by providing insights into the metabolic changes
associated with AD [@jia2021]. This article introduces the
**AlzheiMarkers** R package, which facilitates the analysis and
evaluation of metabolomic biomarkers for distinguishing Alzheimer’s
Disease (AD) patients from Healthy Controls (HC). The package includes
functions for data preparation, model training, evaluation, and
visualization of biomarker importance and model performance through ROC
curves.

## Keywords

Alzheimer’s Disease, Metabolomics, Biomarkers, Machine Learning, R
Package, Random Forest, XGBoost

## 1. Introduction

Alzheimer’s disease (AD) is the most common cause of dementia,
accounting for an estimated 60-80% of cases. Despite extensive research,
AD diagnosis still relies largely on clinical evaluation and
neuroimaging, often identifying the disease at a relatively advanced
stage. Metabolomics, the comprehensive study of metabolites in
biological samples, has emerged as a novel tool for biomarker discovery
in AD. Numerous studies suggest that metabolic dysregulation plays a
critical role in AD pathogenesis.

The **AlzheiMarkers** R package was developed to support the analysis of
metabolomic data for the discovery of potential biomarkers of
Alzheimer’s disease. This package allows users to efficiently load and
preprocess datasets, split data into training and testing sets, train
machine learning models, and visualize model performance. It integrates
widely used machine learning algorithms to distinguish AD patients from
Healthy Controls (HC).

## 2. Materials and Methods

### 2.1 Package Overview

The **AlzheiMarkers** package contains multiple functions that
streamline the workflow from data preparation to model evaluation and
visualization. The primary goals of the package are to facilitate
metabolomic biomarker analysis and enable the exploration of
classification models for differentiating between AD and HC samples.

### 2.2 Data Preparation

The first step involves data preparation, where raw data are imported
and separated into two groups: AD patients and HC. The `prepare_data()`
function reads in a CSV file and returns the full dataset, alongside
separated subsets for AD and HC.

```{r prepare_data, eval=FALSE}
prepare_data <- function(file_path) {
  data <- data.table::fread(file_path)
  data <- as.data.frame(data)
  HC <- dplyr::filter(data, Cohort == "HC")
  AD <- dplyr::filter(data, Cohort == "AD")
  
  list(
    data = data,
    HC = HC,
    AD = AD
  )
}
```

### 2.3 Data Splitting

After the data is prepared, it is split into training and test sets
using the `split_data()` function. This function provides flexibility in
defining the ratio of the dataset that will be used for training.

### 2.4 Model Training

The **AlzheiMarkers** package supports the training of multiple machine
learning models, including Lasso regression, Random Forest, Naive Bayes,
Partial Least Squares (PLS), and XGBoost. Each model is trained using
cross-validation with user-specified parameters. The `train_models()`
function integrates the `caret` package, a powerful tool for building
and tuning machine learning models [@kuhn2008].

```{r train_models, eval=FALSE}
train_models <- function(train_data, n_cv = 10, n_repeat = 10) {
  set.seed(123)
  control <- caret::trainControl(
    method = "repeatedcv",
    number = n_cv,
    repeats = n_repeat,
    classProbs = TRUE,
    summaryFunction = caret::twoClassSummary
  )
  
  predictors <- train_data %>% dplyr::select(-Cohort)
  response <- train_data$Cohort
  
  models <- list(
    lasso = caret::train(x = predictors, y = response,
                         method = "glmnet", metric = "ROC", 
                         trControl = control),
    rf = caret::train(x = predictors, y = response, 
                      method = "rf", metric = "ROC", 
                      trControl = control),
    nb = caret::train(x = predictors, y = response, 
                      method = "naive_bayes", metric = "ROC", 
                      trControl = control),
    pls = caret::train(x = predictors, y = response, 
                       method = "pls", metric = "ROC", 
                       trControl = control),
    xgb = caret::train(x = predictors, y = response, 
                       method = "xgbTree", metric = "ROC", 
                       trControl = control)
  )
  
  models
}
```

### 2.5 Model Evaluation and Visualization

Once the models are trained, their performance can be evaluated using
the `evaluate_models()` function, which provides various evaluation
metrics such as Area Under the ROC Curve (AUC). In addition, the package
includes the `plot_roc_curves()` function to visualize the ROC curves
for each model and `plot_variable_importance()` to plot the most
important variables driving the predictions.

```{r plot+roc_curves, eval=FALSE}
plot_roc_curves <- function(eval_results) {
  ggplot2::ggplot(eval_results$roc, ggplot2::aes(x = 1 - Specificity, 
                                                 y = Sensitivity, 
                                                 color = Model)) +
    ggplot2::geom_line() +
    ggplot2::geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
    ggplot2::theme_minimal() +
    ggplot2::labs(title = "ROC Curves for Different Models")
}
```

### 2.6 Package Vignette

A vignette is provided with the package to guide users through a typical
analysis workflow, including data preparation, model training, and
visualization.

## 3. Results

Using the **AlzheimerBiomarkers** package, users can efficiently train
and evaluate multiple machine learning models to identify potential
biomarkers. By visualizing ROC curves, users can determine which models
provide the highest diagnostic accuracy for distinguishing AD from HC
samples. The variable importance plots offer insights into the most
influential biomarkers.

## 4. Discussion

The **AlzheimerBiomarkers** package provides a user-friendly interface
for conducting biomarker discovery and machine learning-based
classification in the context of Alzheimer's Disease. While the package
supports several common machine learning methods, it can be further
expanded to include additional techniques and functionalities such as
feature selection, ensemble modeling, and deep learning approaches.

## 5. Conclusion

The **AlzheimerBiomarkers** R package is a comprehensive tool for
analyzing metabolomic biomarkers and differentiating AD patients from
healthy controls using machine learning techniques. By providing
functions for data preparation, model training, evaluation, and
visualization, it simplifies the biomarker discovery process in
Alzheimer's research.

## 6. References
